---
title: "Tidyr"
output: html_document
---

# Load Packages
```{r}
library(tidyverse)
library(janitor) 
library(babynames)
```

# Toy Data 
```{r}
cases <- tribble(
  ~Country, ~"2011", ~"2012", ~"2013",
  "FR",       7000,    6900,   7000,
  "DE",       5800,    6000,   6200,
  "US",       15000,   14000, 13000
)

pollution <- tribble(
       ~city, ~size, ~amount,
  "New York", "large",      23,
  "New York", "small",      14,
    "London", "large",      22,
    "London", "small",      16,
   "Beijing", "large",     121,
   "Beijing", "small",     121
)

wide_fail <- tribble(
  ~school, ~teacher,  ~subject,  ~average,
    "UPC",        "A",        "MAT",     .90,
    "UCC",        "B",        "MAT",     .99,
    "UCC",        "A",        "ELA",     .92,
    "UCC",        "B",        "ELA",     .91  
)

# To avoid a distracting detail during class
data(who)
who <- who %>% rename_all(~stringr::str_replace(., "newrel", "new_rel"))
```





# Tidy Data Example
```{r}
population
```

# Toy Data - Wide
```{r}
cases
```


# Your Turn
Draw how the cases data set would look with three columns
country, year, n


```{r}
cases
```


# Practice pivot_longer

```{r}

```


# Toy Data - Long

```{r}
pollution
```

Work out pollution into the collumns,
city, large and small
```{r}

```




# Your Turn - baby names

Reshape the layout of this data. Calculate the percent of male (or female) children by year. Then plot the percent over time. (You can use excel or Tableau for the plot)

```{r}
babynames %>%
  group_by(year, sex) %>% 
  summarize(n = sum(n))
```

***


# Take Aways

Data comes in many formats but R prefers just one: _tidy data_.

A data set is tidy if and only if:

1. Every variable is in its own column
2. Every observation is in its own row
3. Every value is in its own cell (which follows from the above)

What is a variable and an observation may depend on your immediate goal.
